---
title: "STAT220 Oblig 4-5"
author: "Sigbjørn Fjelland"
date: "10/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library in use:
```{r}
library(tinytex)
library(matrixcalc)
library(shape)
library(diagram)
```

\underline{Problem 6.1}

Vi har gitt transition matrix P:
```{r}
library(matrixcalc)
P=matrix(c(0.7, 0.2, 0.1, 0.0, 0.6, 0.4, 0.5, 0.0, 0.5), nrow=3, ncol=3, byrow = TRUE)
dimnames(P)=list(c('0','1','2'), c('0','1','2'))
print(P)
```

a)
  Since $P_{i,j}^{m} = P(x_{m+n}=j|x_{n})$ and $P^{(m)}=P^{m}$:
  
  $\Rightarrow{P^{(3)}=P^{3}}$ og vi får følgende matrise

```{r}
P_3 = matrix.power(P,3)
print(P_3)
print(P_3[1,2])
```
slik at:
$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.008}}$


samme egenskap gjelder også for $P^{(3)}=P^{3}$
```{r}
P_4 = matrix.power(P,4)
print(P_4)
print(P_4[1,2])
```

og vi får da:

$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.0016}}$

b) 
 As we se from the graphic presentation below, it is possible to go from any state to another trough one or more steps:
 
 
```{r}
plotmat(P, box.size = 0.028, arr.lcol='blue')
```

from state 0: 
$0\rightarrow 2\rightarrow 1$

from state 1:
$1\rightarrow 0\rightarrow 2$

from state 2:()
$2\rightarrow 1\rightarrow 0$ and $1\rightarrow 0$
  
It is therfore irreducible.

Periodicity:
    Since the period of a state is the largest $d$ that satisfy following properties: 
      - $p_{ii}^{(n)} = 0$ whenever $n$ is not deviceble by $d$. 
      - The period $i$ is shown by $d(i)$. 
      - If $P_{ii}^{(n)} = 0$, for all $n>0\rightarrow{i}=\infty$
      
      and...
      $i$ is periodic if $d(i)>1$ and aperiodic if $di=1$
      
    It is aperiodic since there are several sequences of steps to go from a state and back again ($i\rightarrow i$),icluding the fact that all the states are self periodic the markov chain is aperiodic.
  
Transient or recurrent? $\rightarrow$ It is recurrent!
  It might stop and loop at a position or between state 2 and 0 wich are comunicating, but eventually it will occure in any of the states by a certanty of 100% as in the formal definition:
   $f_{ii}=P(X_{n} = i, \textrm{for some } n>\geq|w_{0}=i)$ for any state $i$ is Recurrent if $f_{ii}=1$ and transient if Recurrent if $f_{ii}<1$.
  

  