---
title: "STAT220 Oblig 4-5"
author: "Sigbjørn Fjelland"
date: "10/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library in use:
```{r}
library(tinytex)
library(matrixcalc)
library(shape)
library(diagram)
```

\underline{Problem 6.1}

Vi har gitt transition matrix P:
```{r}
library(matrixcalc)
P=matrix(c(0.7, 0.2, 0.1, 0.0, 0.6, 0.4, 0.5, 0.0, 0.5), nrow=3, ncol=3, byrow = TRUE)
dimnames(P)=list(c('0','1','2'), c('0','1','2'))
print(P)
```

a)
  Since $P_{i,j}^{m} = P(x_{m+n}=j|x_{n})$ and $P^{(m)}=P^{m}$:
  
  $\Rightarrow{P^{(3)}=P^{3}}$ og vi får følgende matrise

```{r}
P_3 = matrix.power(P,3)
print(P_3)
print(P_3[1,2])
```
slik at:
$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.008}}$


samme egenskap gjelder også for $P^{(3)}=P^{3}$
```{r}
P_4 = matrix.power(P,4)
print(P_4)
print(P_4[1,2])
```

og vi får da:

$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.0016}}$

b) 
 As we se from the graphic presentation below, it is possible to go from any state to another trough one or more steps:
 
 
```{r}
plotmat(P, box.size = 0.028, arr.lcol='blue')
```

from state 0: 
$0\rightarrow 2\rightarrow 1$

from state 1:
$1\rightarrow 0\rightarrow 2$

from state 2:
$2\rightarrow 1\rightarrow 0$ and $1\rightarrow 0$
  
It is therfore irreducible.


  