---
title: "STAT220 Oblig 4-5"
author: "Sigbjørn Fjelland"
date: "10/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library in use:
```{r}
library(tinytex)
library(matrixcalc)
library(shape)
library(diagram)
library(igraph)
library(devtools)
library(expm)
```
\newpage
\underline{\textbf{Problem 6.1}}

Vi har gitt transition matrix P:
```{r}
P=matrix(c(0.7, 0.2, 0.1, 0.0, 0.6, 0.4, 0.5, 0.0, 0.5), nrow=3, ncol=3, byrow = TRUE)
dimnames(P)=list(c('0','1','2'), c('0','1','2'))
print(P)
```

a)
  Since $P_{i,j}^{m} = P(x_{m+n}=j|x_{n})$ and $P^{(m)}=P^{m}$:
  
  $\Rightarrow{P^{(3)}=P^{3}}$ og vi får følgende matrise

```{r}
P_3 = matrix.power(P,3)
print(P_3)
print(P_3[1,2])
```
slik at:
$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.264}}$


samme egenskap gjelder også for $P^{(3)}=P^{3}$
```{r}
P_4 = matrix.power(P,4)
print(P_4)
print(P_4[1,2])
```

og vi får da:

$P(x_{3}=1|x_{0}=0) = \underline{\underline{0.254}}$

b) 
 As we se from the graphic presentation below, it is possible to go from any state to another trough one or more steps:
 
 
```{r}
h <- graph_from_adjacency_matrix(P, weighted = "prob")
E(h)$prob <- ifelse(is.nan(E(h)$prob), NA, E(h)$prob)
plot(h, edge.label = round(E(h)$prob, 2), edge.arrow.size = .25, edge.curved=-0.2, edge.label.cex = .5)
```

from state 0: 
$0\rightarrow 2\rightarrow 1$

from state 1:
$1\rightarrow 0\rightarrow 2$

from state 2:()
$2\rightarrow 1\rightarrow 0$ and $1\rightarrow 0$
  
It is therfore irreducible.

Periodicity:
    Since the period of a state is the largest $d$ that satisfy following properties: 
      - $p_{ii}^{(n)} = 0$ whenever $n$ is not deviceble by $d$. 
      - The period $i$ is shown by $d(i)$. 
      - If $P_{ii}^{(n)} = 0$, for all $n>0\rightarrow{i}=\infty$
      
      and...
      
  $i$ is periodic if $d(i)>1$ and aperiodic if $di=1$
      
  It is aperiodic since there are several sequences of steps to go from a state and back again ($i\rightarrow i$),
  icluding the fact that all the states are self periodic the markov chain is aperiodic.
  
Transient or recurrent? $\rightarrow$ It is recurrent!
  It might stop and loop at a position or between state 2 and 0 wich are comunicating, but eventually it will occure in any of the states by a certanty of 100% as in the formal definition:
   $f_{ii}=P(X_{n} = i, \textrm{for some } n\geq|w_{0}=i)$ for any state $i$ is Recurrent if $f_{ii}=1$ and transient if Recurrent if $f_{ii}<1$.


\underline{Problem 6.?}

a)
$G: Green$
$R: Red$

We start initiatly at $GR$ and pick a random ball with $p=1/2$. We ar then sent to a state of $RR$ or $GG$ where the probability of being returned to State $GR$ is $100\%$

```{r}
A=matrix(c(0, 1, 0, 0.5, 0, 0.5, 0, 1, 0), nrow=3, ncol=3, byrow = TRUE)
dimnames(A)=list(c('GG','GR','RR'), c('GG','GR','RR'))
```

```{r}
f <- graph_from_adjacency_matrix(A, weighted = "prob")
E(f)$prob <- ifelse(is.nan(E(f)$prob), NA, E(f)$prob)
plot(f, edge.label = round(E(f)$prob, 2),vertex.size=30, edge.arrow.size = .25, edge.curved=0.3, edge.label.cex = .5)
```

Wich gives following matrix


```{r}
print(A)
```

and following for picking red ball in the future entering the state of $GR\rightarrow$ $P(X_{n}=j|X_{0}=1):$ #her må jeg få sendt ett spørsmål
```{r}
print(A[2,])
```


  b)
    $d(GR)=2$ og $d(GG)=d(RR)=4$ som gir største felles nevner 2 og dermed er den periodisk med periode 2.
\newpage

\underline{\textbf{Problem 6.2}}
    Due to previous use of P as variable, it is here substituted for B
  a)
```{r}
B=round(matrix(c((1/3), 0.0, (1/3), 0.0, 0.0, (1/3), 
           (1/2), (1/4), (1/4), 0.0, 0.0, 0.0, 
           0.0, 0.0, (1/2), 0.0, (1/2), 0.0,
           (1/4), (1/4), (1/4), 0.0, 0.0, (1/4),
           0.0, 0.0, 1.0, 0.0, 0.0, 0.0, 
           0.0, 0.0, 0.0, 0.0, 0.0, 1.0), nrow=6, ncol=6, byrow = TRUE),2)
dimnames(B)=list(c('0','1','2','3','4','5'), c('0','1','2','3','4','5'))
print(B)
```


```{r}
g <- graph_from_adjacency_matrix(B, weighted = "prob")
E(g)$prob <- ifelse(is.nan(E(g)$prob), NA, E(g)$prob)
plot(g, edge.label = round(E(g)$prob, 2), edge.arrow.size = .25, edge.label.cex = .5)
```

\newpage
  b)
    \underline{As we se from the graph above:}
  
  We have three recurrent states, 2, 4 and 5, devided in two classes.       
  class 1 - State 5 is self-recurrent
  class 2 - State 2 and 4
  
 
 \textbf{Transient States:}
  
  State 0, 1, and 3 will at some pint possibly loop, but ultimatly it will end up in 
  recurrent state class 1 or 2.


c)
    \underline{Communicating States:}
  
  $(2\leftrightarrow{4})$, $(1\leftrightarrow{1})$, $(2\leftrightarrow{2})$ and $(5\leftrightarrow{5})$
  
  d) Matrix realocated to statespace = {1,3,0,2,4,5}: 
      (Matrix denoted - C to separate matrixes in the code)
```{r}
C=round(matrix(c((1/4), 0.0, (1/2), (1/4), 0.0, 0.0, 
           (1/4), 0.0, (1/4), (1/4), 0.0, (1/4), 
           0.0, 0.0, (1/3), (1/3), 0.0, (1/3),
           0.0, 0.0, 0.0, (1/2), (1/2), 0.0,
           0.0, 0.0, 0.0, 1.0, 0.0, 0.0, 
           0.0, 0.0, 0.0, 0.0, 0.0, 1.0), nrow=6, ncol=6, byrow = TRUE),2)
dimnames(C)=list(c('1','3','0','2','4','5'), c('1','3','0','2','4','5'))

print(C)
```



  e) 
```{r}
R<-round(matrix(c(C[1, 4:6],
C[2, 4:6],
C[3, 4:6]), nrow=3, ncol=3, byrow = TRUE), 2)
dimnames(R)=list(c('1','3','0'), c('2','4','5'))
print(R)
```

This gives $P_{0}(X_{T}=5)=P(X_{T}=5|X_{T-1}=0)=0.3333=\frac{1}{3}$
\newpage

and the expectation:\newline
$E(X_{T}=5|X_{T-1})=1+\sum_{j\in{S}}E(T_{P}=j|X_{T-1}=0)\cdot{P(X_{T}=j|X_{T-1}=0})\\=1+\sum_{j\in{S}}m_{0j}\cdot{P(X_{T}=j|X_{T-1}=0})$\newline

We are proceeding for state 5 such that $m_{j,0}=1$, $m_{2,0}=0$.\newline

$E(X_{T}=5|X_{T-1})=1+m_{0,0}\cdot{P(X_{T}=0|X_{T-1}=0)}\\+m_{2,0}\cdot{P(X_{T}=2|X_{T-1}=0)}+m_{5,0}\cdot{P(X_{T}=5|X_{T-1}=0)}\\=1+1\cdot{\frac{1}{3}}+1\cdot{\frac{1}{3}}+0\cdot{\frac{1}{3}}\\=\underline{\underline{\frac{5}{3}\approx1,667}}$

  f)
for $p^{\infty}$:
```{r}
P_infinity = round(matrix.power(B,1000000),2)
print(P_infinity)
```
C=matrix(c((1/4), 0.0, (1/2), (1/4), 0.0, 0.0, 
           0.0, 0.0, 0.0, (1/2), (1/2), 0.0, 
           0.0, 0.0, (1/3), (1/3), 0.0, (1/3),
           0.0, 0.0, 0.0, (1/2), (1/2), 0.0,
           0.0, 0.0, 0.0, 1.0, 0.0, 0.0, 
           0.0, 0.0, 0.0, 0.0, 0.0, 1.0), nrow=6, ncol=6, byrow = TRUE)
#dimnames(B)=list(c('1','3','0','2','4','5'), c('1','3','0','2','4','5'))
We see the matrix converges in the recurrent states in class 1 and 2.
\newpage

\underline{\textbf{problem 6.3}}

```{r}
mP=round(matrix( c(0.62, 0.00, 0.38, 0.0,
                 0.55, 0.00, 0.45, 0.00,
                 0.00, 0.24, 0.00, 0.7,
                 0.00, 0.19, 0.00, 0.81),nrow=4, ncol=4, byrow = TRUE),2)
dimnames(mP)=list(c('0','1','2','3'), c('0','1','2','3'))
nu<-t(mP[1,])
 # The first row vector of mP
mI<-diag(rep(1,4))
 # unitmatrix of order 4
s<-mI[,1]
 # first unit vector, e_1, in R**4
msnu<-s%*%nu
mR<- mP-msnu

```

Vi har matrisen $P$:
```{r}
print(mP)
```

Vektor $\nu$:
```{r}
print(nu)
```
Vektor $s$:
```{r}
print(s)
```
$s\otimes{\nu}$:
```{r}
print(msnu)
```
matrix $R=P-s\otimes{\nu}$:
```{r}
mR<-mP-msnu
print(mR)
```

b)
  We chose an $M>1$ and set M = 5:
```{r}
M<-5
Gsv <- matrix(0,4,4)

for(n in 0:M){Gsv<-Gsv + mR%^%n}

print(Gsv)
